Evaluating FAIR Digital Object as a distributed object system

### 2023-06-12 latex/main.tex
we have a conclusion!

It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations.

@@

When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used. Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.

@@

The FAIR and Linked Data communities likewise need to recognize the need for simpler, more pragmatic approaches that make it easier for FAIR practitioners to adapt the technologies with "just enough" semantics. We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model. However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available. This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes. Likewise, DOIP implementations like Cordra typically already use HTTP APIs \autocite{DOIPAPIHTTPa}, which can be augmented with signposting headers without necessarily moving to a Linked Data metadata model.

@@

@@

This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreement numbers H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreement numbers HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s  \emph{Horizon Europe funding guarantee} grant numbers 10038963 (EuroScienceGateway), 10038930 (BioDT).

@@

\defbibheading{fdobibliography}{\subsection*{FDO Specifications}}
\printshorthands[heading=fdobibliography]

>>>>

It is worth pointing out at that, except for the DOIP endorsement, all of these documents are conceptual, in the sense that they permit any technical implementation of FDO, if used according to the recommendations. See bibliography \vref{sec:fdo-bibliography} for the citation per document above.

@@

When evaluating FDO implementations against the FDO guidelines (\vref*{sec:doip-fdo-compare}) we see that several technical pieces and community practices still need to be developed and further defined, for instance the FDO type system, how to declare FDO actions, how to resolve persistent identifiers, or how to know which pattern of FDO composition is used. Achieving fully interoperable FAIR digital objects would require further convergence on implementation practices, and it is not given that his need to diverge from the established Web architecture.  It is not clear from FDO guidelines if moving from HTTP/DNS to DOIP/Handle as a way to expose distributed digital objects will benefit FAIR practitioners, when both approaches require additional restrictions, equably implementable, such as using persistent identifiers or pre-defining an object's type.

@@

The FAIR and Linked Data communities likewise need to recognize the need for simpler, more pragmatic approaches that make it easier for FAIR practitioners to adapt the technologies with "just enough" semantics. We have previously proposed the combination of RO-Crate \autocite{10.3233/ds-210053} and Signposting \autocite{vandesompelFAIRSignpostingProfile2022} as a mean to implement FDO \autocite{10.3897/rio.8.e93937} over HTTP using a common Linked Data metadata model. However it may be sufficient to use HTTP-based FAIR Signposting alone to achieve the above list, if one considers only a small metadata model, and rather reference from the signposting which metadata resources are additionally available. This will allow any Linked Data resource to gradually participate in the FDO ecosystem, with minimal effort and non-intrusive implementation changes. FDO implementations like Cordra typically already use HTTP APIs that align with DOIP \autocite{DOIPAPIHTTPa}, these can be augmented with signposting headers without necessarily moving to a Linked Data metadata model.

@@

In this work we have considered FAIR Digital Objects (FDO) as a potential distributed object system and compared FDO with established Web approaches like Linked Data. We have described the background of the Semantic Web and FAIR Digital Objects, and evaluated both using multiple conceptual frameworks. We find that both directions can significantly benefit from each-other and should be aligned further. Namely Linked Data proponents need to make their technologies more approachable, agreeing on predictable and consistent implementations of FAIR principles. The FDO recommendations show that FAIR thinking in this regard need to move beyond data publishing and into machine actionability across digital objects, and with broader community consensus. While flexibility is a necessary ingredient alongside rigidity, the FDO community likewise need to settle on directly implementable specifications rather than just guidelines, and avoid making similar mistakes as the early Semantic Web adopters.

@@

This work was funded by the European Union programmes \emph{Horizon 2020} under grant agreements H2020-INFRAEDI-02-2018 823830 (BioExcel-2), H2020-INFRAEOSC-2018-2 824087 (EOSC-Life) and \emph{Horizon Europe} under grant agreements HORIZON-INFRA-2021-EMERGENCY-01 101046203 (BY-COVID), HORIZON-INFRA-2021-EOSC-01 101057388 (EuroScienceGateway), HORIZON-INFRA-2021-EOSC-01-05 101057344 (FAIR-IMPACT), HORIZON-INFRA-2021-TECH-01 101057437 (BioDT); and by UK Research and Innovation (UKRI) under the UK government’s \emph{Horizon Europe funding guarantee} grants 10038963 (EuroScienceGateway), 10038992(FAIR-IMPACT), 10038930 (BioDT).

@@

\defbibheading{fdobibliography}{\subsection*{FDO Specifications}\label{sec:fdo-bibliography}}
\printshorthands[heading=fdobibliography]
